\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-11 analysis/paper/paper.tex

\Address{
    Sahir Bhatnagar *\\
  McGill Univeristy\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  University of Manitoba\\
  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
  E-mail: \email{max.turgeon@umanitoba.ca}\\
  URL: \url{http://maxturgeon.ca/}\\~\\
      Jesse Islam\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{jesse.islam@mail.mcgill.ca}\\

      James Hanley\\
  McGill Univeristy\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{james.hanley@mcgill.ca}\\
  URL: \url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\~\\
      Olli Saarela\\
  University of Toronto\\
  Dalla Lana School of Public Health, 155 College Street, 6th floor,
  Toronto, Ontario M5T 3M7, Canada\\
  E-mail: \email{olli.saarela@utoronto.ca}\\
  URL: \url{http://individual.utoronto.ca/osaarela/}\\~\\
  }

@@

\hypertarget{literature-review}{%
\section{Literature review}\label{literature-review}}

To conduct our review on current methods available in survival analysis,
a scraper was coded to examine all packages in the survival task view on
CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
of which are not relevant to the problems casebase addresses. To filter
these packages, a scraper was used to search for competing risks,
cumulative incidence functions, non- proportional hazards and
penalization. This reduced our set of 259 packages down to 60. From
here, we manually examined each package to determine which ones are
comparable to casebase, with either a different approach to the same
problems or future features we would like to add to casebase. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models, as casebase is
currently focused on absolute-risk, competing events and single event
analysis.

These packages can belong to three main categories: non-proportional
hazards, regularization and competing risks. Besides these three main
categories, there are extra features of interest as well. Table
\ref{tab:surv-pkgs} summarizes the findings.

Many packages exist that implement semi-parametric Cox models. Of those,
\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package} include penalization, but they implement
the extended cox model permitting intuitive non proportional hazard
modeling while the previous four do not. In terms of penalization, the
benefits of \pkg{casebase} comes from its flexibility. Linear models
have a wide tool set, considerably larger than those made for Cox
regression. \pkg{casebase} permits many regularization tools developed
for linear regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are \pkg{CFC}
\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
\pkg{survival} contain implementations for users to supply any
distribution they would like \citeyearpar{survival-package},
\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
and \pkg{rstpm2} can use splines, permitting a flexible fit over time
\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
\pkg{casebase} is limited to three distributions at the moment, it also
uses splines allowing for distributions that may not be consistent over
survival time.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
differentiating factor between these packages and \pkg{casebase} is that
\pkg{casebase} handles competing risks through multiple logistic
regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
handle interval censoring, whereas \pkg{casebase} currently cannot.
Interval censoring is a simple feature to implement and is a potential
improvement to add to \pkg{casebase}.

Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
\pkg{survival} have implementations for cumulative incidence
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
\pkg{casebase} is parametric, the baseline hazard is known and requires
integration to retrieve the cumulative incidence. This is similar to the
approaches of \pkg{CFC} and \pkg{flexsurv}
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
other packages use semi-parametric approaches to retrieve the baseline
after fitting \citeyearpar{scheike2014estimating},
\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
for their respective parametric and semi-parametric model
implementations \citeyearpar{clements_liu},
\citeyearpar{survival-package}.

@@

Popular survival analysis methods, like Kaplan-Meier and Cox regression,
rely on the notion of risk-set sampling. Case-base sampling can be seen
as an alternative.

As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
expanded in Saarela \citeyearpar{saarela2016case}), writing the
likelihood arising from this data-generating mechanism using the
framework of non-homogenous Poisson processes, we eventually reach an
expression where each person-moment's contribution is of the form
\[\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},\] where \(N(t)\) is
the counting process associated with the event of interest,
\(\lambda(t)\) is the corresponding hazard function, and \(\rho(t)\) is
the hazard function for the Poisson process associated with case-base
sampling. This parametric form suggests that we can readily estimate
log-hazards of the form \(\log(\lambda(t)) = g(t; X)\) using logistic
regression, where each observation corresponds to a person moment, the
function \(g(t; X)\) is linear in a finite number of parameters, and
where we treat \(\log(\rho(t))\) as an offset.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest performing case-base samping \emph{uniformly}, i.e.~to sample
the base series uniformly from the study base. In terms of Poisson
processes, this sampling strategy corresponds to a time-homogeneous
Poisson process with intensity equal to \(b/B\), where \(b\) is the
number of sampled observations in the base series, and \(B\) is the
total population-time for the study base. More complex examples are also
available; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the probabilities of the sampling
mechanism are proportional to the cardiovascular disease event rate
given by the Framingham score.

\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
(number of units) of a point in prognostic/prospective time and \(x\) is
the realization of the vector \(X\) of variates based on the patient's
profile and intervention (if any). Different functions of \(t\) lead to
different parametric hazard models.

The simplest of these models is the one-parameter exponential
distribution which is obtained by taking the hazard function to be
constant over the range of \(t\).

\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]

The instantaneous failure rate is independent of \(t\), so that the
conditional chance of failure in a time interval of specified length is
the same regardless of how long the individual has been on study a.k.a
the memoryless property (Kalbfleisch and Prentice, 2002).

\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]

Use of \(log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]

@@

Finally, the hazard function is fitted to the data using the function
\code{glm}, unless there is more than one type event (i.e.~in a
competing-risk analysis), in which case it uses the multinomical
regression capabilities of the \pkg{VGAM} package. This package was
selected for its ability to fit multinomial regression models with an
offset.

Once a model-fit object has been returned by \code{fitSmoothHazard}, all
the familiar summary and diagnostic functions are available:
\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
package provides one more functionality: it computes risk functions from
the model fit. For the case of a single event, it uses the familiar
identity \[S(t) = \exp\left(-\int_0^t h(u;X) du\right).\] The integral
is computed using either the \code{stats::integrate} function or
Monte-Carlo integration. For the case of a competing-event analysis, the
event-specific risk is computed using a nested double integral; in this
setting, Monte-Carlo integration is faster and more flexible than
\code{stats::integrate}. This is due to the fact that the computation
involves a double loop over the selected time points; Monte-Carlo
integration performs this step by using the vectorized \code{rowSums}
and \code{colSums} functions.

To decide between a single-event and a competing-event analysis, we
created \code{absoluteRisk} as an \code{S3} generic, with methods for
both \code{glm} and \code{CompRisk} objects (the latter inherits from
\code{vglm} as well). The method dispatch system of \proglang{R} then
takes care of matching the correct input to the correct methodology,
without the user's intervention.

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
R>                                     data = ERSPC, 
R>                                     ratio = 100)
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #demonstrating incidence density after imputation
R> plot(casebase::popTime(as.data.frame(completeData),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
R> #to determine the hazard function
R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
R>                              formula_time = ~d.time,alpha=1,ratio=100)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
R> #hazard for cox using glmnet
R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> plot(coxglmFit)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxFit$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet
R> newDataCox=newData[nonzero_covariate_cox$X1-1]
R> 
R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # A plot to compare all three Absolute risk (Commulative Incidence)
R> plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
R> lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
R> legend("bottom", 
R>        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
R>        col = c("black","red","Blue"),
R>        lty = c(1, 1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

>>>>

\Address{
    Sahir Bhatnagar *\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  University of Manitoba\\
  186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
  E-mail: \email{max.turgeon@umanitoba.ca}\\
  URL: \url{https://maxturgeon.ca/}\\~\\
      Jesse Islam\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{jesse.islam@mail.mcgill.ca}\\

      James Hanley\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{james.hanley@mcgill.ca}\\
  URL: \url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\~\\
      Olli Saarela\\
  University of Toronto\\
  Dalla Lana School of Public Health, 155 College Street, 6th floor,
  Toronto, Ontario M5T 3M7, Canada\\
  E-mail: \email{olli.saarela@utoronto.ca}\\
  URL: \url{http://individual.utoronto.ca/osaarela/}\\~\\
  }

@@

As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
expanded in Saarela \citeyearpar{saarela2016case}), writing the
likelihood arising from this data-generating mechanism using the
framework of non-homogeneous Poisson processes, we eventually reach an
expression where each person-moment's contribution is of the form
\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
process associated with the event of interest, \(h(t)\) is the
corresponding hazard function, and \(\rho(t)\) is the hazard function
for the Poisson process associated with case-base sampling. This
parametric form suggests that we can readily estimate log-hazards of the
form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
observation corresponds to a person moment, the function \(g(t; X)\) is
linear in a finite number of parameters, and where we treat
\(-\log(\rho(t))\) as an offset.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
the base series uniformly from the study base. In terms of Poisson
processes, this sampling strategy corresponds essentially to a
time-homogeneous Poisson process with intensity equal to \(b/B\), where
\(b\) is the number of sampled observations in the base series, and
\(B\) is the total population-time for the study base. More complex
examples are also available; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the probabilities of the sampling
mechanism are proportional to the cardiovascular disease event rate
given by the Framingham score.

\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
individual's covariate profile. Different functions of \(t\) lead to
different parametric hazard models. The simplest of these models is the
one-parameter exponential distribution which is obtained by taking the
hazard function to be constant over the range of \(t\).

\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]

The instantaneous failure rate is independent of \(t\), so that the
conditional chance of failure in a time interval of specified length is
the same regardless of how long the individual has been in the study.
This is also known as the \emph{memoryless property} (Kalbfleisch and
Prentice, 2002).

\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]

Use of \(\log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]

For competing-risk analyses with \(J\) possible events, we can show that
each person-moment's contribution of the likelihood is of the form

\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
be expected, this functional form is similar to the terms appearing in
the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}

\hypertarget{existing-packages}{%
\section{Existing packages}\label{existing-packages}}

Survival analysis is an important branch of applied statistics and
epidemiology. Accordingly, there is a vast ecosystem of \code{R}
packages implementing different methods. In this section, we describe
how the functionalities of \pkg{casebase} compare to these packages.

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with a description
of some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: \pkg{CoxRidge}
\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
\pkg{survival} \citeyearpar{survival-package}. On the other hand,
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, this is the only package to offer
this functionality.
\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks. The differentiating factor between these packages and
\pkg{casebase} is that \pkg{casebase} handles competing risks through
multiple logistic regression.
\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

Finally, several packages include functions to estimate the cumulative
incidence function. The corresponding methods generally fall into two
categories: transformation of the estimated hazard function, and
semi-parametric estimation of the baseline hazard. The first category
broadly corresponds to parametric survival models, where the full hazard
is explictly modeled. Using this estimate, the survival function and the
cumulative incidence function can be obtained using their functional
relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv},
\pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows
this approach for both single-event and competing-event analyses. The
second category outlined above broadly corresponds to Cox models. These
models do not model the full hazard function, and therefore the baseline
hazard needs to be estimated separately in order to estimate the
survival function. This is achieved using semi-parametric estimators
(e.g.~Breslow's estimator). Packages that implement this approach
include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As
mentioned in the introduction, a key distinguishing factor between these
two approaches is that the first category leads to smooth estimates of
the cumulative incidence function, whereas the second category produces
estimates in the form of step-wise functions. This was one of the main
motivations for introducting case-base sampling in survival analysis.

@@

@@

To fit the hazard function, we provide several approaches that are
available via the \code{family} parameter. These approaches are:

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using Lasso
  or elastic-net. This functionality is provided through the
  \code{glmnet} package \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \code{gam} package
  \citep{hastie1987generalized}.
\item
  \code{gbm}: This option provides support for \emph{Gradient Boosted
  Trees} via the \code{gbm} package. This feature is still experimental.
\end{itemize}

In the case of multiple events, the hazard is fitted via multinomial
regression as performed by the \pkg{VGAM} package. This package was
selected for its ability to fit multinomial regression models with an
offset.

Once a model-fit object has been returned by \code{fitSmoothHazard}, all
the familiar summary and diagnostic functions are available:
\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
package provides one more functionality: it computes risk functions from
the model fit. For the case of a single event, it uses the familiar
identity \begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t h(u;X) du\right).
\end{equation} The integral is computed using either the
\code{stats::integrate} function or Monte-Carlo integration. The risk
function (or cumulative incidence function) is then defined as
\begin{equation}\label{eqn:CI}
CI(t) = 1 - S(t).
\end{equation}

For the case of a competing-event analysis, the event-specific risk is
computed using the following procedure: first, we compute the overall
survival function (i.e.~for all event types):

\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
From this, we can derive the event-specific subdensities:

\[ f_j(t) = h_j(t)S(t).\]

Finally, by integrating these subdensities, we obtain the event-specific
cumulative incidence functions:

\[ CI_j(t) = \int_0^t f_j(u)du.\]

We created \code{absoluteRisk} as an \code{S3} generic, with methods for
the different types of outputs of \code{fitSmoothHazard}. The method
dispatch system of \proglang{R} then takes care of matching the correct
output to the correct methodology for calculating the cumulative
incidence function, without the user's intervention.

In the following sections, we illustrate these functionalities in the
context of three case studies.

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
R>                                     data = ERSPC, 
R>                                     ratio = 100)
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #demonstrating incidence density after imputation
R> plot(casebase::popTime(as.data.frame(completeData),time = "d.time",event = "death"))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{../figures/poptimePostImputation-1}

\caption{\label{fig:ptPostI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study.}\label{fig:poptimePostImputation}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
R> #to determine the hazard function
R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
R+                              formula_time = ~d.time,alpha=1,ratio=100)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
R> #hazard for cox using glmnet
R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> plot(coxglmFit)
\end{CodeInput}

\begin{center}\includegraphics{../figures/coxHazAbsolute-1} \end{center}

\begin{CodeInput}
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxFit$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet
R> newDataCox=newData[nonzero_covariate_cox$X1-1]
R> 
R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # A plot to compare all three Absolute risk (Commulative Incidence)
R> plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
R> lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
R> legend("bottom", 
R+        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
R+        col = c("black","red","Blue"),
R+        lty = c(1, 1, 1), 
R+        bg = "gray90")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{../figures/abSupportComparison-1}

\caption{\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue.}\label{fig:abSupportComparison}
\end{figure}
\end{CodeChunk}

\hypertarget{case-study-4stanford-heart-transplant-data}{%
\section{Case study 4--Stanford Heart Transplant
Data}\label{case-study-4stanford-heart-transplant-data}}

Although the previous case studies provide a broad overview of the
capabilities of \pkg{casebase}, they all have two properties in common:

\begin{itemize}
  \item All covariates are fixed, i.e. they do not change over time;
  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
\end{itemize}

Case-base sampling has been used in the literature to study vaccination
safety, where the exposure period was defined as the week following
vaccination \citep{saarela2015case}. Hence, the main covariate of
interest, i.e.~exposure to the vaccine, was changing over time. In this
context, case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

The purpose of the next case study is to show how \pkg{casebase} can
also be used for survival analysis problems that do not share the
properties above. To this end, we will use the Stanford Heart Transplant
Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
use the version available in the \pkg{survival} package.

Recall the setting of this study: patients were admitted to the Stanford
program after meeting with their physician and determining that they
were unlikely to respond to other forms of treatment. After enrollment,
the program searches for a suitable donor for the patient, which can
take anywhere between a few days to almost a year. We are interested in
the effect of a heart transplant on survival; therefore, the patient is
considered exposed only after the transplant has occurred.

As before, we can look at the population-time plot for a graphical
summary of the event incidence. As we can see, most events occur early
during the follow-up period, and therefore we do not expect the hazard
to be constant.

\begin{CodeInput}
R> library(survival)
R> library(casebase)
R> 
R> stanford_popTime <- popTime(jasa, time = "futime", 
R+                             event = "fustat")
R> plot(stanford_popTime)
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}

Since the exposure is time-dependent, we need to manually define the
exposure variable \emph{after} case-base sampling and \emph{before}
fitting the hazard function. For this reason, we will use the
\texttt{sampleCaseBase} function directly.

\begin{CodeInput}
R> library(tidyverse)
R> library(lubridate)
R> 
R> cb_data <- sampleCaseBase(jasa, time = "futime", 
R+                           event = "fustat", ratio = 10)
\end{CodeInput}
\end{CodeChunk}

Next, we will compute the number of days from acceptance into the
program to transplant, and we use this variable to determine whether
each population-moment is exposed or not.

\begin{CodeInput}
R> # Define exposure variable
R> cb_data <- mutate(cb_data,
R+                   txtime = time_length(accept.dt %--% tx.date, 
R+                                        unit = "days"),
R+                   exposure = case_when(
R+                     is.na(txtime) ~ 0L,
R+                     txtime > futime ~ 0L,
R+                     txtime <= futime ~ 1L
R+                   ))
\end{CodeInput}
\end{CodeChunk}

Finally, we can fit the hazard using various linear predictors.

\begin{CodeInput}
R> library(splines)
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R+                         data = cb_data, time = "futime")
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R+                         data = cb_data, time = "futime")
R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
R+                         data = cb_data, time = "futime")
R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
R+                         data = cb_data, time = "futime")
\end{CodeInput}
\end{CodeChunk}

Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
term between exposure and follow-up time. In other words, this model no
longer exhibit proportional hazards. The evidence of non-proportionality
of hazards in the Stanford Heart Transplant data has been widely
discussed \citep{arjas1988graphical}.

We can then compare the goodness of fit of these four models using the
Akaike Information Criterion (AIC).

\begin{CodeInput}
R> # Compute AIC
R> c("Model1" = AIC(fit1),
R+   "Model2" = AIC(fit2),
R+   "Model3" = AIC(fit3),
R+   "Model4" = AIC(fit4))
\end{CodeInput}

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    803    763    731    723
\end{CodeOutput}
\end{CodeChunk}

As we can, the best fit is the fourth model. By visualizing the hazard
functions for both exposed and unexposed individuals, we can more
clearly see how the hazards are no longer proportional.

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(exposure = c(0, 1),
R+                            futime = seq(0, 1000,
R+                                         length.out = 100))
R> # Set the offset to zero
R> hazard_data$offset <- 0 
R> # Use predict to get the fitted values, and exponentiate to 
R> # transform to the right scale
R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
R+                                  type = "link"))
R> # Add labels for plots
R> hazard_data$Status = factor(hazard_data$exposure,
R+                             labels = c("NoTrans", "Trans"))
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R+     geom_line() +
R+     theme_minimal() +
R+     theme(legend.position = 'top') +
R+     ylab('Hazard') + xlab('Follow-up time')
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}

The non-proportionality seems to be more pronounced at the beginning of
follow-up than the end. Finally, we can turn these estimate of the
hazard function into estimates of the cumulative incidence functions.
